Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-08 tex/body.tex
Minor

For each of these 245 benchmarks, we considered up to three randomly selected
mutated forms of its conjecture $\psi$. In particular, we considered unique
conjectures that are obtained as a result of swapping a subterm of $\psi$ at
one position with another of the same type at another position.
Note that benchmarks created in this way have a high likelihood of having small
and easy-to-find countermodels. In total, we considered 213 mutated forms of
conjectures from \isa and 427 mutated forms of conjectures from \leon. We will
call these sets \isam and \leonm, respectively.

Thus, our benchmark set consist of 885 benchmarks (640 mutants plus the
original 245). We considered these 885 benchmarks both before and after the
translation~$\conv$. % as described in Section~\ref{sec:encoding}.
For example, \isa contains 79 original benchmarks $\varphi$ and 79 translated
benchmarks $\conv(\varphi)$.

@@

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples for formulas involving recursive
function definitions.
For convenience, it
has been implemented in the development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written using the $\definefunreccmd$ command
from the upcoming SMT-LIB 2.5 standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as %follows:
%
\begin{quote}
%\small -- no small, unless you convince me that these mixtures of font sizes
% are aesthetic and serve a purpose --jb
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\cvc adds the annotated quantified formula
\begin{longv}
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
\end{longv}
\begin{shortv}
$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$
to its list of assertions,
which after rewriting becomes
$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$
\end{shortv}
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the translation $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. 
For the example above, it outputs
%\ajr{prefer having the full output like this here?}
%\jb{I like it the way you did it now, but I took the freedom to reformat the output slightly
%(to avoid the small font---I generally try to avoid ad hoc font mixtures---and to show
%the nesting more clearly).}
%it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.
%
\begin{quote}
%\small
\begin{verbatim}
(model
  (define-fun s (($x1 Int)) Int 
     (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
        (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
           (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
              (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
                 (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
  (define-fun c () Int 14)
)
\end{verbatim}
\end{quote}

>>>>

For each of these 245 benchmarks, we considered up to three randomly selected
mutated forms of its conjecture $\psi$. In particular, we considered unique
conjectures that are obtained as a result of swapping a subterm of $\psi$ at
one position with another of the same type at another position.
Note that benchmarks created in this way have a high likelihood of having small
and easy-to-find countermodels. In total, we considered 213 mutated forms of
conjectures from \isa and 427 mutated forms of conjectures from \leon. We will
call these sets \isam and \leonm, respectively.
Thus, our benchmark set consist of 885 benchmarks (640 mutants plus the
original 245). We considered these 885 benchmarks both before and after the
translation~$\conv$. % as described in Section~\ref{sec:encoding}.
For example, \isa contains 79 original benchmarks $\varphi$ and 79 translated
benchmarks $\conv(\varphi)$.

@@

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples for formulas involving recursive
function definitions.
For convenience, it
has been implemented in the development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written using the $\definefunreccmd$ command
from the upcoming SMT-LIB 2.5 standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as %follows:
%
\begin{quote}
%\small -- no small, unless you convince me that these mixtures of font sizes
% are aesthetic and serve a purpose --jb
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\begin{longv}
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
\end{longv}
\begin{shortv}
\cvc adds
$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$
to its list of assertions,
which after rewriting becomes
$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$.\end{shortv}
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the translation $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. 
For the example above, it outputs
%\ajr{prefer having the full output like this here?}
%\jb{I like it the way you did it now, but I took the freedom to reformat the output slightly
%(to avoid the small font---I generally try to avoid ad hoc font mixtures---and to show
%the nesting more clearly).}
%it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.
%
\begin{quote}
%\small
\begin{verbatim}
(model
  (define-fun s (($x1 Int)) Int 
     (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
        (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
           (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
              (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
                 (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
  (define-fun c () Int 14)
)
\end{verbatim}
\end{quote}
